Estimating the number of clusters using Cross Validation

### 2015-06-06 cvclust.tex
Add moment lemma

@@

Given a sample $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to
$\{ Y_i \}_{i=1}^{n}$.  If $k = 1$ and $n$ is large enough, then the centroid
will be close to $\E(Y) = 0$, in the sense that the centroid converges almost
surely to this value as $n$ tends to infinity \citep{pollard1981strong}.
Similarly, if $k = 2$ and $n$ is large enough, then the centroids will tend to
$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  In
$k = 2$ case, then if $n$ is large enough, the cluster assignment will be
determined according to whether $Y > 0$.

>>>>

\section{Technical Lemmas}

\begin{lemma}\label{lem:truncated-normal-moments}

If $Z$ is a standard normal random variable, then
\[
  \E(Z \mid a < Z < b)
    = - \frac{\varphi(b) - \varphi(a)}
             {\Phi(b) - \Phi(a)}
\]
and
\[
  \E\{(Z - \delta)^2 \mid a < Z < b\}
    = \delta^2 + 1
    - \frac{  (b - 2 \delta) \varphi(b)
            - (a - 2 \delta) \varphi(a)}
           {\Phi(b) - \Phi(a)}
\]
for all constants $a$, $b$, and $\delta$, where
$\phi(z)$ and $\Phi(z)$ are the standard normal probability density and
cumulative distribution functions.  These expressions are valid
for $a = -\infty$ or $b = \infty$ by taking limits.

\end{lemma}
\begin{proof}
We will derive the expression for the second moment.
Integrate by parts to get
\begin{align*}
  \E[ (Z - \delta)^2 1\{Z < b\}]
    &= \int_{-\infty}^b (z - \delta)^2 \varphi(z) \, dz \\
    &= (\delta^2 + 1) \Phi(b) - (b - 2 \delta) \varphi(b).
\end{align*}
Now,
\[
  \E\{(Z - \delta)^2 \mid a < Z < b\}
    =
    \frac{  \E[ (Z - \delta)^2 1\{Z < b\}]
          - \E[ (Z - \delta)^2 1\{Z < a\}]}
         { \Phi(b) - \Phi(a) }.
\]
\end{proof}

Lemma~\ref{lem:truncated-normal-moments} has some important special cases:
\begin{align*}
  \E\{Z \mid Z > 0\} &= 2 \varphi(0) = \sqrt{2 / \pi}, \\
  \E\{(Z - \delta)^2 \mid Z > 0 \}
    &= \delta^2 + 1 - 4 \delta \varphi(0), \\
  \E\{(Z - \delta)^2 \mid Z < 0 \}
    &= \delta^2 + 1 + 4 \delta \varphi(0).
\end{align*}

@@

Given a sample $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to
$\{ Y_i \}_{i=1}^{n}$.  If $k = 1$ and $n$ is large enough, then the centroid
will be close to $\E(Y) = 0$, in the sense that the centroid converges almost
surely to this value as $n$ tends to infinity \citep{pollard1981strong}.
Similarly, if $k = 2$ and $n$ is large enough, then the centroids will tend to
$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  We
have used Lemma~\ref{lem:truncated-normal-moments} to compute the
expectations.  In $k = 2$ case, then if $n$ is large enough, the cluster
assignment will be determined according to whether $Y > 0$.
